---
title: "Basics of R Answers"
output: html_notebook
---

#Answers to questions from Basics_of_R.Rmd

1. Can you successfully set your directory to open a file? Try to open the additional file I have included in Canvas (species.txt) and create a dataframe with the file. Name your new dataframe "Heliconius".
```{r}
setwd("/Users/jessicajudson/Downloads/") #an example directory to set
getwd() #prints your working directory (where you are in your computer)
#the dataset we want to download is a text file (the ending of the filename is .txt), so we use the read.table command. Let's name the file "Heliconius". **IMPORTANT** if you name the object "data", R overwrites your old variable called "data". Make sure you don't overwrite your variables!
Heliconius <- read.table("species.txt", header = TRUE, sep = "\t")
```
To make sure it downloaded correctly, open it by clicking it in the environment to the right.

2. Can you successfully download the packages we need for the first computational activity? The packages are ape (should have been installed above) and pegas
```{r}
install.packages("ape")
install.packages("pegas")
```

3. When you want to load a package for use in your current R Studio session, what should you do? Do this for pegas.
```{r}
library(pegas)
```

4. Can you look up how to do things in a package? For example, can you figure out what the seg.sites command does in ape, and how to use it?
```{r}
?seg.sites
```

