---
title: "Chapter 4.1.1 Computational Work"
output: html_notebook
---

#Chapter 4.1.1: Levels of diversity maintained by a balance between mutation and drift

###Preparation:
1. Set your working directory: 
```{r}

```

2. Load plotrix to your library. 
```{r}
#install.packages("plotrix")
library(plotrix)
```

3. Reload the simulate.pop function into your R functions for use:
```{r}
simulate.pop<-function(N.vec=rep(5,30), const.RS=TRUE,  mutation= TRUE, mut.rate=  0.1, for.class= TRUE, initial.state="all.black",plot.freqs=FALSE,mult.pop=FALSE,pops=FALSE){
   #  c(rep(10,5),rep(3,2),rep(10,5),rep(3,2),rep(10,5))  #
	stopifnot(initial.state %in% c("all.black","all.diff","two.alleles","single.mut") )

	if(plot.freqs){ layout(c(1,2)); par(mar=c(1,2,0,1))}
	if(for.class){
		line.lwd<-1
		line.col<-"black"
		mut.line.lwd<-1
		mut.line.col<-"black"
	
	}else{
		line.lwd<-0.5
		line.col<-"grey"
		mut.line.lwd<-1
		mut.line.col<-"grey"
	}

	num.gens<- length(N.vec)-1	

	if(!mult.pop){
		ind.pop.par<-matrix(1,nrow=max(N.vec),ncol=num.gens+1)
		ind.pop<-matrix(1,nrow=max(N.vec),ncol=num.gens+1)
	}else{
			ind.pop.par<-pops[["ind.pop.par"]]
			ind.pop<-pops[["ind.pop"]]
	}
	
	num.gens<- length(N.vec)-1
	offset<-0.1
	plot(c(1,num.gens),c(0.5,max(N.vec))+c(-offset,offset),type="n",axes=FALSE,xlab="",ylab="")
	mtext(side=1,line=0,"Generations")
	 text(1,0.5,"Past")
	 text(num.gens-1,0.5,"Present")
	 
	 track.cols<- list()
	N <-N.vec[1]
	 if(initial.state=="all.black") my.cols<-rep("black",2*N)  #sample(rainbow(2*N))
	 if(initial.state=="all.diff") my.cols<-sample(rainbow(2*N))
	 if(initial.state=="two.alleles")  my.cols<-  rep(c("blue","red"),N)
	 if(initial.state=="single.mut")  my.cols<-  c("red",rep("blue",2*N-1))
	 stopifnot((2*N)==length(my.cols))
	
	 track.cols[[1]]<-my.cols
	points(rep(1,N),1:N+offset, pch=19,cex=1.3,col=my.cols[(1:N)*2])
	 points(rep(1,N),1:N-offset, pch=19,cex=1.3,col=my.cols[(1:N)*2-1])
	
	for(i in 1:num.gens){
			
		N.new<-N.vec[i+1]
		N.old<-N.vec[i]
		points(rep(i,N.old),1:N.old+offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2])
		points(rep(i,N.old),1:N.old-offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2-1])

		 new.cols<-rep("black",2*N.new)
		
		if(const.RS){ 
			repro.success<-rep(1/N.old,N.old)
		}else{
			repro.success<-sample(c(rep(0.5/(N.old),N.old-2),c(0.25,0.25)),replace=FALSE)
			}
	
		 for(ind in 1:N.new){

				this.pop.par <- ind.pop.par[ind,i+1]
				available.pars <- (1:N.old)[which(ind.pop[1:N.old,i] == this.pop.par)]
				par<-sample(available.pars,2,replace=FALSE,prob=repro.success[which(ind.pop[1:N.old,i] == this.pop.par)])
		
				which.allele.1<-sample(c(-1,1),1)
				if(i != num.gens){ lines(c(i,i+1), c(par[1]+which.allele.1*offset,ind-offset),col=line.col,lwd=line.lwd)}
				new.cols[2*ind-1]<- my.cols[2*par[1] +ifelse(which.allele.1==1,0,-1)]
		
				which.allele.2<-sample(c(-1,1),1)
				if(i != num.gens){ lines(c(i,i+1), c(par[2]+which.allele.2*offset,ind+offset),col=line.col,lwd=line.lwd)}
				new.cols[2*ind]<- my.cols[2*par[2] +ifelse(which.allele.2==1,0,-1)]
		
				if(mutation){
					if(runif(1)<mut.rate){ 
							new.cols[2*ind-1]<- sample(rainbow(4*N),1)
							if(i != num.gens){ lines(c(i,i+1), c(par[1]+which.allele.1*offset,ind-offset),col=mut.line.col,lwd=mut.line.lwd)}
		
						}
					if(runif(1)<mut.rate){ 
						new.cols[2*ind]<- sample(rainbow(4*N),1)
						if(i != num.gens){ lines(c(i,i+1), c(par[2]+which.allele.2*offset,ind+offset),col=mut.line.col,lwd=mut.line.lwd)}
				} 
				
			}
		}	
		##redraw points to cover lines		 
		points(rep(i,N.old),1:N.old+offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2])
		points(rep(i,N.old),1:N.old-offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2-1])

		 my.cols<-new.cols
		 track.cols[[i+1]]<-my.cols
		if(!const.RS) sapply(which(repro.success>1/N.old), function(ind){ draw.circle(x=i,y=ind,radius=0.2,nv=100,border=NULL,col=NA,lty=1,lwd=1)})
	}
#	recover()
	if(plot.freqs){
		plot(c(1,num.gens),c(0,1),type="n",axes=FALSE,xlab="",ylab="")
		all.my.cols<-unique(unlist(track.cols))
		
		if(!mult.pop){ 
			my.col.freqs<-sapply(track.cols,function(my.gen){sapply(all.my.cols,function(my.col){sum(my.gen==my.col)})})

			sapply(all.my.cols,function(col.name){lines(my.col.freqs[col.name,]/(2*N.vec),col=col.name,lwd=2)});
			}else{
			
			for(pop in 1:max(ind.pop)){
				my.col.freqs<-sapply(1:num.gens, function(gen){
	#			recover()
					my.gen<-track.cols[[gen]]
					if(all(ind.pop.par[ind.pop[,gen]==pop,gen]==0)) return(rep(NA,length(all.my.cols)))  #if pop doesn't exist in this gen.
				
					these.inds<-which(ind.pop[,gen]==pop)
					my.gen<-c(my.gen[these.inds*2],my.gen[these.inds*2-1])
					sapply(all.my.cols,function(my.col){
						sum(my.gen==my.col)
					})})
				rownames(my.col.freqs)<-		all.my.cols
				sapply(all.my.cols[-length(all.my.cols)],function(col.name){lines(my.col.freqs[col.name,]/(2*5),col=col.name,lwd=2,lty=pop)});	
			}
			}
		
		axis(2)
	}
}

```

#Let's begin by simulating the concept of Mutation-Drift Balance (Fig. 4.7):
```{r}
#simulate a population of 5 with mutations at a rate of 0.1 (figure 4.7)
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation= TRUE, mut.rate=  0.1, for.class= TRUE,plot.freqs=TRUE,initial.state="all.black") #note that we plot the changing frequencies with the plot.freqs=TRUE option.
```

In this simulation, both drift and mutation are acting to change allele frequencies over time. As the mutation rate in this simulation is set very high (0.1), the levels of polymorphism are maintained even as drift acts to remove certain alleles from the population. In other words, drift removes variation, while mutation creates it. 

#What happens if you increase the mutation rate? Try 0.2:
```{r}
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.2, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
#increased mutation: increased rare alleles in the population, increased heterozygosity
```

#What happens when you decrease the mutation rate? (in humans, average mutation rate is 1.5x10^-8 per base) Let's try 0.01:
```{r}
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
#decreased mutation: drift acts to decrease heterozygosity faster on average than mutation produces heterozygosity
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
```

#Let's try a scenario where we take 5 genetically diverse individuals (all heterozygotes for different alleles) and run the simulation with this lower mutation rate of 0.01 (which is still much much higher than is realistic for most organisms)
```{r}
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.diff")
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.diff")
#drift acts to decrease heterozygosity and mutation cannot replenish that genetic diversity
```

Above, we have seen how drift and mutation act to change levels of heterozygosity in a population. However, the component we have not talked about yet, which is population size, is directly predictive of the power of genetic drift (1/2N is the rate of decay of heterozygosity) and should thus be considered when we discuss mutation-drift balance. Additionally, population size influences the population-scaled mutation rate θ, which is calculated as:

θ=4Nμ

where μ is the mutation rate at a given locus, and N is the population size. As population size increases, θ also increases. Let's test this with our simulations:
```{r}
simulate.pop(N.vec=rep(50,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
simulate.pop(N.vec=rep(50,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
#compare to smaller population size
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
simulate.pop(N.vec=rep(5,30), const.RS=TRUE,  mutation=TRUE, mut.rate=  0.01, for.class= TRUE, plot.freqs = TRUE, initial.state="all.black")
```

#Big picture: Small populations are extremely vulnerable to loss of genetic variation, as drift is stronger AND population-scaled mutation rate is lower! And "species with larger population sizes should have proportionally higher levels of neutral polymorphism" (Coop 2020)