---
title: "Chapter 4.1.2 Computational Work"
output: html_notebook
---

#Chapter 4.1.2: The effective population size

###Preparation:
1. Set your working directory: 
```{r}

```

2. Load plotrix to your library. 
```{r}
#install.packages("plotrix")
library(plotrix)
```

3. Reload the simulate.pop function into your R functions for use:
```{r}
simulate.pop<-function(N.vec=rep(5,30), const.RS=TRUE,  mutation= TRUE, mut.rate=  0.1, for.class= TRUE, initial.state="all.black",plot.freqs=FALSE,mult.pop=FALSE,pops=FALSE){
   #  c(rep(10,5),rep(3,2),rep(10,5),rep(3,2),rep(10,5))  #
	stopifnot(initial.state %in% c("all.black","all.diff","two.alleles","single.mut") )

	if(plot.freqs){ layout(c(1,2)); par(mar=c(1,2,0,1))}
	if(for.class){
		line.lwd<-1
		line.col<-"black"
		mut.line.lwd<-1
		mut.line.col<-"black"
	
	}else{
		line.lwd<-0.5
		line.col<-"grey"
		mut.line.lwd<-1
		mut.line.col<-"grey"
	}

	num.gens<- length(N.vec)-1	

	if(!mult.pop){
		ind.pop.par<-matrix(1,nrow=max(N.vec),ncol=num.gens+1)
		ind.pop<-matrix(1,nrow=max(N.vec),ncol=num.gens+1)
	}else{
			ind.pop.par<-pops[["ind.pop.par"]]
			ind.pop<-pops[["ind.pop"]]
	}
	
	num.gens<- length(N.vec)-1
	offset<-0.1
	plot(c(1,num.gens),c(0.5,max(N.vec))+c(-offset,offset),type="n",axes=FALSE,xlab="",ylab="")
	mtext(side=1,line=0,"Generations")
	 text(1,0.5,"Past")
	 text(num.gens-1,0.5,"Present")
	 
	 track.cols<- list()
	N <-N.vec[1]
	 if(initial.state=="all.black") my.cols<-rep("black",2*N)  #sample(rainbow(2*N))
	 if(initial.state=="all.diff") my.cols<-sample(rainbow(2*N))
	 if(initial.state=="two.alleles")  my.cols<-  rep(c("blue","red"),N)
	 if(initial.state=="single.mut")  my.cols<-  c("red",rep("blue",2*N-1))
	 stopifnot((2*N)==length(my.cols))
	
	 track.cols[[1]]<-my.cols
	points(rep(1,N),1:N+offset, pch=19,cex=1.3,col=my.cols[(1:N)*2])
	 points(rep(1,N),1:N-offset, pch=19,cex=1.3,col=my.cols[(1:N)*2-1])
	
	for(i in 1:num.gens){
			
		N.new<-N.vec[i+1]
		N.old<-N.vec[i]
		points(rep(i,N.old),1:N.old+offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2])
		points(rep(i,N.old),1:N.old-offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2-1])

		 new.cols<-rep("black",2*N.new)
		
		if(const.RS){ 
			repro.success<-rep(1/N.old,N.old)
		}else{
			repro.success<-sample(c(rep(0.5/(N.old),N.old-2),c(0.25,0.25)),replace=FALSE)
			}
	
		 for(ind in 1:N.new){

				this.pop.par <- ind.pop.par[ind,i+1]
				available.pars <- (1:N.old)[which(ind.pop[1:N.old,i] == this.pop.par)]
				par<-sample(available.pars,2,replace=FALSE,prob=repro.success[which(ind.pop[1:N.old,i] == this.pop.par)])
		
				which.allele.1<-sample(c(-1,1),1)
				if(i != num.gens){ lines(c(i,i+1), c(par[1]+which.allele.1*offset,ind-offset),col=line.col,lwd=line.lwd)}
				new.cols[2*ind-1]<- my.cols[2*par[1] +ifelse(which.allele.1==1,0,-1)]
		
				which.allele.2<-sample(c(-1,1),1)
				if(i != num.gens){ lines(c(i,i+1), c(par[2]+which.allele.2*offset,ind+offset),col=line.col,lwd=line.lwd)}
				new.cols[2*ind]<- my.cols[2*par[2] +ifelse(which.allele.2==1,0,-1)]
		
				if(mutation){
					if(runif(1)<mut.rate){ 
							new.cols[2*ind-1]<- sample(rainbow(4*N),1)
							if(i != num.gens){ lines(c(i,i+1), c(par[1]+which.allele.1*offset,ind-offset),col=mut.line.col,lwd=mut.line.lwd)}
		
						}
					if(runif(1)<mut.rate){ 
						new.cols[2*ind]<- sample(rainbow(4*N),1)
						if(i != num.gens){ lines(c(i,i+1), c(par[2]+which.allele.2*offset,ind+offset),col=mut.line.col,lwd=mut.line.lwd)}
				} 
				
			}
		}	
		##redraw points to cover lines		 
		points(rep(i,N.old),1:N.old+offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2])
		points(rep(i,N.old),1:N.old-offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2-1])

		 my.cols<-new.cols
		 track.cols[[i+1]]<-my.cols
		if(!const.RS) sapply(which(repro.success>1/N.old), function(ind){ draw.circle(x=i,y=ind,radius=0.2,nv=100,border=NULL,col=NA,lty=1,lwd=1)})
	}
#	recover()
	if(plot.freqs){
		plot(c(1,num.gens),c(0,1),type="n",axes=FALSE,xlab="",ylab="")
		all.my.cols<-unique(unlist(track.cols))
		
		if(!mult.pop){ 
			my.col.freqs<-sapply(track.cols,function(my.gen){sapply(all.my.cols,function(my.col){sum(my.gen==my.col)})})

			sapply(all.my.cols,function(col.name){lines(my.col.freqs[col.name,]/(2*N.vec),col=col.name,lwd=2)});
			}else{
			
			for(pop in 1:max(ind.pop)){
				my.col.freqs<-sapply(1:num.gens, function(gen){
	#			recover()
					my.gen<-track.cols[[gen]]
					if(all(ind.pop.par[ind.pop[,gen]==pop,gen]==0)) return(rep(NA,length(all.my.cols)))  #if pop doesn't exist in this gen.
				
					these.inds<-which(ind.pop[,gen]==pop)
					my.gen<-c(my.gen[these.inds*2],my.gen[these.inds*2-1])
					sapply(all.my.cols,function(my.col){
						sum(my.gen==my.col)
					})})
				rownames(my.col.freqs)<-		all.my.cols
				sapply(all.my.cols[-length(all.my.cols)],function(col.name){lines(my.col.freqs[col.name,]/(2*5),col=col.name,lwd=2,lty=pop)});	
			}
			}
		
		axis(2)
	}
}

```

In the last computational activity, you saw through simulations that population size dictates drift and population-scaled mutation rate θ. While this is true, our population size variable has been very simplified in these simulations to assume that every individual has an equal opportunity to pass on their alleles. 

That isn't very realistic in nature. It is easy to imagine that in many species with social hierarchies, not every individual gets to mate and produce offspring. In other species, perhaps certain individuals are isolated and never find mates. To accommodate for this, we use the term `effective population size` to better represent the actual number of actively reproducing individuals in a population.

#To demonstrate the influence of effective population size on genetic drift, let's use our simulate.pop function.
```{r}
#initial simulation for comparison:
simulate.pop(N.vec=rep(10,19), const.RS=TRUE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
#simulate a repeated loss of individuals or a lack of reproduction of certain individuals
repeated.crash<- c(rep(10,5),rep(3,2),rep(10,5),rep(3,2),rep(10,5))
#create a scenario like that in Fig. 4.8
simulate.pop(N.vec=repeated.crash, const.RS=TRUE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
simulate.pop(N.vec=repeated.crash, const.RS=TRUE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
simulate.pop(N.vec=repeated.crash, const.RS=TRUE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
```
In the above scenario, there are two time points when only six individuals get to pass their alleles on to the next generation, which consists of only three individuals. With each subsequent reduction in individuals, called a bottleneck, more genetic variation is lost due to drift (the random loss of alleles). While the population size overall is 10 individuals at the end of the simulation, the genetic diversity is more reflective of the three individuals that the population was reduced to at those two time points. 

#Variance in reproductive success influences effective population size, which in turn influences drift (Fig. 4.9)
```{r}
#simulation for comparison
	simulate.pop(N.vec=rep(10,10), const.RS=TRUE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")

#circled individuals have increased reproductive success
	simulate.pop(N.vec=rep(10,10), const.RS=FALSE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
	simulate.pop(N.vec=rep(10,10), const.RS=FALSE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
	simulate.pop(N.vec=rep(10,10), const.RS=FALSE,  mutation= FALSE, for.class= TRUE, initial.state="all.diff")
	
#As you can see, variation can be lost much faster due to drift, as certain individuals have greater reproductive success and pass their alleles on at a higher rate, while other individuals do not and their alleles are lost from drift.
```
