---
title: "Chapter 4.3 Computational Work"
output: html_notebook
---

#Chapter 4.3: The coalescent process of a sample of alleles 
*Note, I said incorrectly there would be a 4.2 activity, but there is not. This is the final computational activity for chapter 4.

###Preparation:
1. Set your working directory: 
```{r}

```

2. Load plotrix to your library. 
```{r}
#install.packages("plotrix")
library(plotrix)
```

3. Reload the simulate.pop function into your R functions for use:
```{r}
simulate.pop<-function(N.vec=rep(5,30), const.RS=TRUE,  mutation= TRUE, mut.rate=  0.1, for.class= TRUE, initial.state="all.black",plot.freqs=FALSE,mult.pop=FALSE,pops=FALSE){
   #  c(rep(10,5),rep(3,2),rep(10,5),rep(3,2),rep(10,5))  #
	stopifnot(initial.state %in% c("all.black","all.diff","two.alleles","single.mut") )

	if(plot.freqs){ layout(c(1,2)); par(mar=c(1,2,0,1))}
	if(for.class){
		line.lwd<-1
		line.col<-"black"
		mut.line.lwd<-1
		mut.line.col<-"black"
	
	}else{
		line.lwd<-0.5
		line.col<-"grey"
		mut.line.lwd<-1
		mut.line.col<-"grey"
	}

	num.gens<- length(N.vec)-1	

	if(!mult.pop){
		ind.pop.par<-matrix(1,nrow=max(N.vec),ncol=num.gens+1)
		ind.pop<-matrix(1,nrow=max(N.vec),ncol=num.gens+1)
	}else{
			ind.pop.par<-pops[["ind.pop.par"]]
			ind.pop<-pops[["ind.pop"]]
	}
	
	num.gens<- length(N.vec)-1
	offset<-0.1
	plot(c(1,num.gens),c(0.5,max(N.vec))+c(-offset,offset),type="n",axes=FALSE,xlab="",ylab="")
	mtext(side=1,line=0,"Generations")
	 text(1,0.5,"Past")
	 text(num.gens-1,0.5,"Present")
	 
	 track.cols<- list()
	N <-N.vec[1]
	 if(initial.state=="all.black") my.cols<-rep("black",2*N)  #sample(rainbow(2*N))
	 if(initial.state=="all.diff") my.cols<-sample(rainbow(2*N))
	 if(initial.state=="two.alleles")  my.cols<-  rep(c("blue","red"),N)
	 if(initial.state=="single.mut")  my.cols<-  c("red",rep("blue",2*N-1))
	 stopifnot((2*N)==length(my.cols))
	
	 track.cols[[1]]<-my.cols
	points(rep(1,N),1:N+offset, pch=19,cex=1.3,col=my.cols[(1:N)*2])
	 points(rep(1,N),1:N-offset, pch=19,cex=1.3,col=my.cols[(1:N)*2-1])
	
	for(i in 1:num.gens){
			
		N.new<-N.vec[i+1]
		N.old<-N.vec[i]
		points(rep(i,N.old),1:N.old+offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2])
		points(rep(i,N.old),1:N.old-offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2-1])

		 new.cols<-rep("black",2*N.new)
		
		if(const.RS){ 
			repro.success<-rep(1/N.old,N.old)
		}else{
			repro.success<-sample(c(rep(0.5/(N.old),N.old-2),c(0.25,0.25)),replace=FALSE)
			}
	
		 for(ind in 1:N.new){

				this.pop.par <- ind.pop.par[ind,i+1]
				available.pars <- (1:N.old)[which(ind.pop[1:N.old,i] == this.pop.par)]
				par<-sample(available.pars,2,replace=FALSE,prob=repro.success[which(ind.pop[1:N.old,i] == this.pop.par)])
		
				which.allele.1<-sample(c(-1,1),1)
				if(i != num.gens){ lines(c(i,i+1), c(par[1]+which.allele.1*offset,ind-offset),col=line.col,lwd=line.lwd)}
				new.cols[2*ind-1]<- my.cols[2*par[1] +ifelse(which.allele.1==1,0,-1)]
		
				which.allele.2<-sample(c(-1,1),1)
				if(i != num.gens){ lines(c(i,i+1), c(par[2]+which.allele.2*offset,ind+offset),col=line.col,lwd=line.lwd)}
				new.cols[2*ind]<- my.cols[2*par[2] +ifelse(which.allele.2==1,0,-1)]
		
				if(mutation){
					if(runif(1)<mut.rate){ 
							new.cols[2*ind-1]<- sample(rainbow(4*N),1)
							if(i != num.gens){ lines(c(i,i+1), c(par[1]+which.allele.1*offset,ind-offset),col=mut.line.col,lwd=mut.line.lwd)}
		
						}
					if(runif(1)<mut.rate){ 
						new.cols[2*ind]<- sample(rainbow(4*N),1)
						if(i != num.gens){ lines(c(i,i+1), c(par[2]+which.allele.2*offset,ind+offset),col=mut.line.col,lwd=mut.line.lwd)}
				} 
				
			}
		}	
		##redraw points to cover lines		 
		points(rep(i,N.old),1:N.old+offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2])
		points(rep(i,N.old),1:N.old-offset, pch=19,cex=1.3,col=my.cols[(1:N.old)*2-1])

		 my.cols<-new.cols
		 track.cols[[i+1]]<-my.cols
		if(!const.RS) sapply(which(repro.success>1/N.old), function(ind){ draw.circle(x=i,y=ind,radius=0.2,nv=100,border=NULL,col=NA,lty=1,lwd=1)})
	}
#	recover()
	if(plot.freqs){
		plot(c(1,num.gens),c(0,1),type="n",axes=FALSE,xlab="",ylab="")
		all.my.cols<-unique(unlist(track.cols))
		
		if(!mult.pop){ 
			my.col.freqs<-sapply(track.cols,function(my.gen){sapply(all.my.cols,function(my.col){sum(my.gen==my.col)})})

			sapply(all.my.cols,function(col.name){lines(my.col.freqs[col.name,]/(2*N.vec),col=col.name,lwd=2)});
			}else{
			
			for(pop in 1:max(ind.pop)){
				my.col.freqs<-sapply(1:num.gens, function(gen){
	#			recover()
					my.gen<-track.cols[[gen]]
					if(all(ind.pop.par[ind.pop[,gen]==pop,gen]==0)) return(rep(NA,length(all.my.cols)))  #if pop doesn't exist in this gen.
				
					these.inds<-which(ind.pop[,gen]==pop)
					my.gen<-c(my.gen[these.inds*2],my.gen[these.inds*2-1])
					sapply(all.my.cols,function(my.col){
						sum(my.gen==my.col)
					})})
				rownames(my.col.freqs)<-		all.my.cols
				sapply(all.my.cols[-length(all.my.cols)],function(col.name){lines(my.col.freqs[col.name,]/(2*5),col=col.name,lwd=2,lty=pop)});	
			}
			}
		
		axis(2)
	}
}

```


#This activity will focus on Chapter 4.3.2: Demography and the coalescent. 

Specifically, let's revisit our bottleneck example in 4.1.2, just simulating a single bottleneck this time and allowing mutations (Fig. 4.19):

```{r}
#Bottleneck: population decreases to a few individuals for a period of time

#Single bottleneck:
single.crash<- c(rep(10,9),rep(3,2),rep(10,6))

#simulate a bottleneck with no mutations
simulate.pop(N.vec=single.crash, const.RS=TRUE,  mutation= TRUE, mut.rate = 0.01, for.class= TRUE, initial.state="all.diff")
simulate.pop(N.vec=single.crash, const.RS=TRUE,  mutation= TRUE, mut.rate = 0.01, for.class= TRUE, initial.state="all.diff")
simulate.pop(N.vec=single.crash, const.RS=TRUE,  mutation= TRUE, mut.rate = 0.01, for.class= TRUE, initial.state="all.diff")

#What happens to diversity?
```
You should see in this simulation that genetic diversity decreases during the bottleneck as drift removes rarer alleles. As this bottleneck is shorter, lineages that "escape" coalescing during the bottleneck coalesce more deeply in time, and there is thus an excess of intermediate frequency alleles. This would result in a positive Tajima's D.

#What if the bottleneck is very strong?
```{r}
single.crash<- c(rep(10,9),rep(2,10),rep(10,5)) 
simulate.pop(N.vec=single.crash, const.RS=TRUE,  mutation=TRUE, mut.rate = 0.01, for.class= TRUE, initial.state="all.black")
simulate.pop(N.vec=single.crash, const.RS=TRUE,  mutation=TRUE, mut.rate = 0.05, for.class= TRUE, initial.state="all.black")
simulate.pop(N.vec=single.crash, const.RS=TRUE,  mutation=TRUE, mut.rate = 0.05, for.class= TRUE, initial.state="all.black")
#generally, should see more rare alleles in the population (negative Tajima's D)
```

#Similar to severe bottlenecks, population growth has a similar outcome with respect to the site frequency spectrum: an excess of rare alleles

```{r}
pop.growth<- c(rep(2,3),2^c(1:4,4))
simulate.pop(N.vec=pop.growth, const.RS=TRUE,  mutation= TRUE, mut.rate=  0.05, for.class= TRUE, initial.state="all.black")
simulate.pop(N.vec=pop.growth, const.RS=TRUE,  mutation= TRUE, mut.rate=  0.05, for.class= TRUE, initial.state="all.black")
simulate.pop(N.vec=pop.growth, const.RS=TRUE,  mutation= TRUE, mut.rate=  0.05, for.class= TRUE, initial.state="all.black")
#should see more rare variants in the population following expansion, similar to a strong bottleneck.
```


