---
title: "Chapter 10.1 Computational Activity"
output: html_notebook
---

#Chapter 10: One-Locus Models of Selection

###Preparation:
1. Set your working directory: 
```{r}
options(scipen=10)
```

2. We don't need any packages for this exercise.

#10.1 Balancing selection and the selective maintenance of polymorphism

##We saw based on a simulation of directional selection that we expect a loss of variation over time as selection acts to remove genotypes with lower fitness from the population. But how do we see a maintenance of phenotypic/genetic variation?

There are three ways:
1. Mutation-drift balance
2. Selection maintains variation
3. Mutation-selection balance

#For this activity, we are going to focus on #2, selection maintaining variation in a population.

#10.1.1 Heterozygote advantage: "heterozygotes are fitter than either of the homozygotes" (Coop 2020)

#Let's add to our simulation of selection to account for our understanding of the relative fitnesses of different genotypes under heterozygote advantage (see the table in your text).
```{r}
number_of_generations = 100 #how many generations do you want the simulation to run?

#relative fitnesses
##here we are setting the relative fitnesses of each of the genotypes, and now we are incorporating the selection coefficients against homozygotes.
s1 = 0.1 #set s, the selection coefficient for the first homozygote
s2 = 0.2 #for the second homozygote
h = 0 #set h, the dominance coefficient

fitness_aa = 1-s1 #changes from 1 to be 1-s1 because the highest relative fitness now belongs to the heterozygote
fitness_ab = 1
fitness_bb = 1-s2

#Starting frequencies of alleles and genotypes
freq_a = 0.1
 
freq_b = 1 - freq_a
freq_aa = freq_a^2
freq_ab = 2 * freq_a * freq_b
freq_bb = freq_b^2
 
#create dataframes for allele and genotype frequencies:
genotypes = data.frame(t = 0, cbind(freq_aa, freq_ab, freq_bb))
alleles_frequency = data.frame(t = 0, cbind(freq_a, freq_b))
 
#Now we will simulate over each subsequent generation. 
for (t in 1 : number_of_generations) {
 
    x_freq_a = freq_aa + 0.5 * freq_ab #get starting freq. of a allele
    x_freq_b = freq_bb + 0.5 * freq_ab #and b allele. Need to redo this for each generation as selection changes allele frequencies.
 
    x_freq_aa = freq_aa * fitness_aa #here you calculate the frequency of a genotype multiplied by its relative fitness, which we need for each genotype to calculate wbar, the mean fitness of the population of zygotes (equation 10.22)
    x_freq_ab = freq_ab * fitness_ab
    x_freq_bb = freq_bb * fitness_bb
     
    wbar = x_freq_aa + x_freq_ab + x_freq_bb #here we add those together as in eq. 10.22 to get wbar
     
    freq_a = ((freq_aa*fitness_aa)+(freq_a*freq_b*fitness_ab))/wbar #see equation 10.26
    freq_b = 1 - freq_a #to get the freq. of the other allele, 1-p
    
    #recalculate genotype frequencies in next generation:
    freq_aa = freq_a^2
    freq_ab = 2 * freq_a * freq_b
    freq_bb = freq_b^2
      
    #add information to table
    genotypes = rbind(genotypes, data.frame(t, cbind(freq_aa, freq_ab, freq_bb)))
    alleles_frequency = rbind(alleles_frequency, data.frame(t, cbind(freq_a, freq_b)))
}
```

#We can look at our resulting fitnesses:
```{r}
fitness_aa #=1-s1
fitness_ab #=1
fitness_bb #=1-s2
```
#You should notice that the heterozygote fitness is greater than either homozygote's fitness, because the heterozygote has the highest relative fitness in the simulation.

#We can plot these changes over time:
```{r}
plot(alleles_frequency$t, alleles_frequency$freq_a,type="l",col="red", xlab="Generations", ylab="Allele Frequency", ylim = c(0,1)) 
lines(alleles_frequency$t, alleles_frequency$freq_b,type="l",col="blue")

plot(genotypes$t, genotypes$freq_aa,type="l",col="red", xlab="Generations", ylab="Genotype Frequency", ylim = c(0,1))
lines(genotypes$t, genotypes$freq_ab,type="l",col="green")
lines(genotypes$t, genotypes$freq_bb,type="l",col="blue")
```
#Given the plots above, how is selection acting on the phenotype to change the genotype frequency? Which allele is decreased due to selection? Instead of directional selection, you see a maintenance of variation in the population.

#From equation 10.39, we can even predict the equilibrium frequency of the a allele.
```{r}
s2/(s1+s2)
```

#Does that match our plot?
```{r}
plot(alleles_frequency$t, alleles_frequency$freq_a,type="l",col="red", xlab="Generations", ylab="Allele Frequency", ylim = c(0,1)) 
lines(alleles_frequency$t, alleles_frequency$freq_b,type="l",col="blue")
abline(0.6666667, 0, col = "red", lty = "dashed")
```

#Underdominance: in this scenario, the heterozygote is less fit than either of the homozygotes.

```{r}
number_of_generations = 100 #how many generations do you want the simulation to run?

#relative fitnesses
##here we are setting the relative fitnesses of each of the genotypes, and now we are incorporating the selection coefficients favoring homozygotes.
s1 = 0.1 #set s, the selection coefficient for the first homozygote
s2 = 0.2 #for the second homozygote
h = 0 #set h, the dominance coefficient

fitness_aa = 1+s1 #changes from 1 to be 1+s1 because there is now a selective advantage of being a homozygote
fitness_ab = 1
fitness_bb = 1+s2

#Starting frequencies of alleles and genotypes
freq_a = 0.1
 
freq_b = 1 - freq_a
freq_aa = freq_a^2
freq_ab = 2 * freq_a * freq_b
freq_bb = freq_b^2
 
#create dataframes for allele and genotype frequencies:
genotypes = data.frame(t = 0, cbind(freq_aa, freq_ab, freq_bb))
alleles_frequency = data.frame(t = 0, cbind(freq_a, freq_b))
 
#Now we will simulate over each subsequent generation. 
for (t in 1 : number_of_generations) {
 
    x_freq_a = freq_aa + 0.5 * freq_ab #get starting freq. of a allele
    x_freq_b = freq_bb + 0.5 * freq_ab #and b allele. Need to redo this for each generation as selection changes allele frequencies.
 
    x_freq_aa = freq_aa * fitness_aa #here you calculate the frequency of a genotype multiplied by its relative fitness, which we need for each genotype to calculate wbar, the mean fitness of the population of zygotes (equation 10.22)
    x_freq_ab = freq_ab * fitness_ab
    x_freq_bb = freq_bb * fitness_bb
     
    wbar = x_freq_aa + x_freq_ab + x_freq_bb #here we add those together as in eq. 10.22 to get wbar
     
    freq_a = ((freq_aa*fitness_aa)+(freq_a*freq_b*fitness_ab))/wbar #see equation 10.26
    freq_b = 1 - freq_a #to get the freq. of the other allele, 1-p
    
    #recalculate genotype frequencies in next generation:
    freq_aa = freq_a^2
    freq_ab = 2 * freq_a * freq_b
    freq_bb = freq_b^2
      
    #add information to table
    genotypes = rbind(genotypes, data.frame(t, cbind(freq_aa, freq_ab, freq_bb)))
    alleles_frequency = rbind(alleles_frequency, data.frame(t, cbind(freq_a, freq_b)))
}
```

#We can look at our resulting fitnesses:
```{r}
fitness_aa #=1+s1
fitness_ab #=1
fitness_bb #=1+s2
```
#And plot the changes over time:
```{r}
plot(alleles_frequency$t, alleles_frequency$freq_a,type="l",col="red", xlab="Generations", ylab="Allele Frequency", ylim = c(0,1)) 
lines(alleles_frequency$t, alleles_frequency$freq_b,type="l",col="blue")

plot(genotypes$t, genotypes$freq_aa,type="l",col="red", xlab="Generations", ylab="Genotype Frequency", ylim = c(0,1))
lines(genotypes$t, genotypes$freq_ab,type="l",col="green")
lines(genotypes$t, genotypes$freq_bb,type="l",col="blue")
```

#What is happening here? What if we change the starting allele frequency?
```{r}
number_of_generations = 100 #how many generations do you want the simulation to run?

#relative fitnesses
##here we are setting the relative fitnesses of each of the genotypes, and now we are incorporating the selection coefficients favoring homozygotes.
s1 = 0.1 #set s, the selection coefficient for the first homozygote
s2 = 0.2 #for the second homozygote
h = 0 #set h, the dominance coefficient

fitness_aa = 1+s1 #changes from 1 to be 1+s1 because there is now a selective advantage of being a homozygote
fitness_ab = 1
fitness_bb = 1+s2

#Starting frequencies of alleles and genotypes
freq_a = 0.7
 
freq_b = 1 - freq_a
freq_aa = freq_a^2
freq_ab = 2 * freq_a * freq_b
freq_bb = freq_b^2
 
#create dataframes for allele and genotype frequencies:
genotypes = data.frame(t = 0, cbind(freq_aa, freq_ab, freq_bb))
alleles_frequency = data.frame(t = 0, cbind(freq_a, freq_b))
 
#Now we will simulate over each subsequent generation. 
for (t in 1 : number_of_generations) {
 
    x_freq_a = freq_aa + 0.5 * freq_ab #get starting freq. of a allele
    x_freq_b = freq_bb + 0.5 * freq_ab #and b allele. Need to redo this for each generation as selection changes allele frequencies.
 
    x_freq_aa = freq_aa * fitness_aa #here you calculate the frequency of a genotype multiplied by its relative fitness, which we need for each genotype to calculate wbar, the mean fitness of the population of zygotes (equation 10.22)
    x_freq_ab = freq_ab * fitness_ab
    x_freq_bb = freq_bb * fitness_bb
     
    wbar = x_freq_aa + x_freq_ab + x_freq_bb #here we add those together as in eq. 10.22 to get wbar
     
    freq_a = ((freq_aa*fitness_aa)+(freq_a*freq_b*fitness_ab))/wbar #see equation 10.26
    freq_b = 1 - freq_a #to get the freq. of the other allele, 1-p
    
    #recalculate genotype frequencies in next generation:
    freq_aa = freq_a^2
    freq_ab = 2 * freq_a * freq_b
    freq_bb = freq_b^2
      
    #add information to table
    genotypes = rbind(genotypes, data.frame(t, cbind(freq_aa, freq_ab, freq_bb)))
    alleles_frequency = rbind(alleles_frequency, data.frame(t, cbind(freq_a, freq_b)))
}
plot(alleles_frequency$t, alleles_frequency$freq_a,type="l",col="red", xlab="Generations", ylab="Allele Frequency", ylim = c(0,1)) 
lines(alleles_frequency$t, alleles_frequency$freq_b,type="l",col="blue")

plot(genotypes$t, genotypes$freq_aa,type="l",col="red", xlab="Generations", ylab="Genotype Frequency", ylim = c(0,1))
lines(genotypes$t, genotypes$freq_ab,type="l",col="green")
lines(genotypes$t, genotypes$freq_bb,type="l",col="blue")
```
#There are two stable equilibria the alleles reach, either a frequency of zero or a frequency of one. The heterozygotes have a fitness disadvantage, thus ultimately the homozygotes will fix in the population, but this requires that one of the two alleles must be lost. There is an equilibrium that can be reached between the two alleles in this situation, similar to the above situation for a heterozygote advantage, but this equilibrium is highly unstable.  
```{r}
#equilibrium:
s2/(s1+s2)
```

#What if we set our initial allele frequency to this value?
```{r}
number_of_generations = 100 #how many generations do you want the simulation to run?

#relative fitnesses
##here we are setting the relative fitnesses of each of the genotypes, and now we are incorporating the selection coefficients favoring homozygotes.
s1 = 0.1 #set s, the selection coefficient for the first homozygote
s2 = 0.2 #for the second homozygote
h = 0 #set h, the dominance coefficient

fitness_aa = 1+s1 #changes from 1 to be 1+s1 because there is now a selective advantage of being a homozygote
fitness_ab = 1
fitness_bb = 1+s2

#Starting frequencies of alleles and genotypes
freq_a = s2/(s1+s2)
 
freq_b = 1 - freq_a
freq_aa = freq_a^2
freq_ab = 2 * freq_a * freq_b
freq_bb = freq_b^2
 
#create dataframes for allele and genotype frequencies:
genotypes = data.frame(t = 0, cbind(freq_aa, freq_ab, freq_bb))
alleles_frequency = data.frame(t = 0, cbind(freq_a, freq_b))
 
#Now we will simulate over each subsequent generation. 
for (t in 1 : number_of_generations) {
 
    x_freq_a = freq_aa + 0.5 * freq_ab #get starting freq. of a allele
    x_freq_b = freq_bb + 0.5 * freq_ab #and b allele. Need to redo this for each generation as selection changes allele frequencies.
 
    x_freq_aa = freq_aa * fitness_aa #here you calculate the frequency of a genotype multiplied by its relative fitness, which we need for each genotype to calculate wbar, the mean fitness of the population of zygotes (equation 10.22)
    x_freq_ab = freq_ab * fitness_ab
    x_freq_bb = freq_bb * fitness_bb
     
    wbar = x_freq_aa + x_freq_ab + x_freq_bb #here we add those together as in eq. 10.22 to get wbar
     
    freq_a = ((freq_aa*fitness_aa)+(freq_a*freq_b*fitness_ab))/wbar #see equation 10.26
    freq_b = 1 - freq_a #to get the freq. of the other allele, 1-p
    
    #recalculate genotype frequencies in next generation:
    freq_aa = freq_a^2
    freq_ab = 2 * freq_a * freq_b
    freq_bb = freq_b^2
      
    #add information to table
    genotypes = rbind(genotypes, data.frame(t, cbind(freq_aa, freq_ab, freq_bb)))
    alleles_frequency = rbind(alleles_frequency, data.frame(t, cbind(freq_a, freq_b)))
}
plot(alleles_frequency$t, alleles_frequency$freq_a,type="l",col="red", xlab="Generations", ylab="Allele Frequency", ylim = c(0,1)) 
lines(alleles_frequency$t, alleles_frequency$freq_b,type="l",col="blue")

plot(genotypes$t, genotypes$freq_aa,type="l",col="red", lty=4, lwd=3, xlab="Generations", ylab="Genotype Frequency", ylim = c(0,1))
lines(genotypes$t, genotypes$freq_ab,type="l",col="green", lty=2, lwd=3)
lines(genotypes$t, genotypes$freq_bb,type="l",col="blue", lwd=3)
```
#Pretty awesome right? As long as there is no disruption from any random occurrences, we can expect that underdominance with a starting allele frequency equal to the equilibrium frequency s2/(s1+s2) will cause the system to remain with the same allele frequencies even under selection. However, the moment the population deviates from that equilibrium frequency (which we know it will eventually due to drift/mutation/etc.), the population rapidly stabilizes to the equilibrium of zero or one. That is why this is called an unstable equilibrium.
