---
title: "Chapter 14 Computational Activity"
output: html_notebook
---

#Chapter 14: Interaction of Multiple Selected Loci

###Preparation:
1. Set your working directory
```{r}

```

2. We will need the package "sf" for this exercise. You may already have this package installed, so I have commented this command out. If the library command works, you shouldn't need to install it.
```{r}
#install.packages("sf")
library(sf)
```

3. Set up the functions to create the figures seen in Coop 2020. These are sourced from Dr. Coop's code that accompanies the text.

```{r}
#Function #1: Simulations

#example values so I can show you what the function is doing:
n.gens<-500
#haplotype frequencies:
p<-  c(0.001,0.099,0,0.9); names(p)<-c("AB","Ab","aB","ab")
#relative fitnesses of each haplotype:
w.add<-c(1,0.95,1,0.95);names(w.add)<-names(p)
w.mat<-outer(w.add,w.add,FUN="+")

#function
two.loc.sims<-function(p,w.mat,r){
  #If the haplotype frequencies don't add to one, stop
	stopifnot(sum(p)==1)
  #create an empty matrix to fill with changing haplotype frequencies
	p.array<-matrix(NA,nrow=n.gens,ncol=4)
	#Create empty matrix for our D values
	d.array<-NA
	#Set our first row to be equal to the initial haplotype frequencies set when you call the function.
	p.array[1,]<-p
	colnames(p.array)<-names(p)
	#Begin our loop to measure changes in haplotype frequencies over time.
	for(gen in 2:n.gens){
	  #Multiply haplotype frequencies by their relative fitnesses. Remember that each diploid individual has two haplotypes, so the number of haplotypes with a beneficial/deleterious allele matters. This is recreating the table you see on p.229
		w.marg<-apply(w.mat,1,function(w.marg){w.marg*p}) 
		#get the sums of the columns of the table to get our marginal haplotype frequencies
		w.marg<-colSums(w.marg)
		#calculate our D, which measures "the excess of coupling to repulsion gametes". In other words, the haplotype associations. "If D = 0 we’ll say the two loci are in linkage equilibrium, while if D > 0 or D < 0 we’ll say that the loci are in linkage disequilibrium"
		D<-p["AB"] - (p["AB"]+p["Ab"])*(p["AB"]+p["aB"])
		d.array<-c(d.array,D)
		#"where the coupling haplotypes 1 and 4 use +D and repulsion haplotypes 2 and 3 use −D"
		D.vec<- c(-D,+D,+D,-D)
		
		#"Our mean fitness wbar is the sum of all the entries in the table"
		wbar<-sum(p*w.marg)
		
		##Eq. 14.3 to calculate new haplotype frequencies
		p.new<- (p*w.marg + r*D.vec*w.mat["AB","ab"])/wbar
		p.array[gen,]<-p.new
		p<-p.new
	}
	return(cbind(p.array,d.array))
}
```

#Function 2, Plotting function:
```{r}
stack.freqs.plot<-function(p.out,my.title=""){
	stacked.freqs<-t(apply(p.out[,c("Ab","AB","ab","aB")],1,cumsum))
	plot(stacked.freqs[,"Ab"],ylim=c(0,1),type="l",xlab="Generations",ylab="Frequencies",cex.lab=1.4,cex.axis=1.2,main=my.title,cex.main=1.4)
	my.x<-1:n.gens
	x.polygon<-c(my.x,rev(my.x))
#	recover()
	polygon(x=x.polygon,c(stacked.freqs[,"Ab"],rep(0,n.gens)),col="blue")
 	pol <-st_polygon(list(cbind(c(x.polygon,1), c(stacked.freqs[,"Ab"],rep(0,n.gens),stacked.freqs[,"Ab"][1]))))
 	centroid<-st_centroid(pol)
	text(centroid[1],centroid[2],col="white","Ab",cex=1.4) 

	polygon(x=x.polygon,c(stacked.freqs[,"AB"],rev(stacked.freqs[,"Ab"])),col="purple")
 	pol <-st_polygon(list(cbind(c(x.polygon,1), c(stacked.freqs[,"AB"],rev(stacked.freqs[,"Ab"]),stacked.freqs[,"AB"][1]))))
 	centroid<-st_centroid(pol)
	text(centroid[1],centroid[2],col="white","AB",cex=1.4) 
	
	polygon(x=x.polygon,c(stacked.freqs[,"AB"],rev(stacked.freqs[,"aB"])),col="white")
	pol <-st_polygon(list(cbind(c(x.polygon,1) ,c(stacked.freqs[,"AB"],rev(stacked.freqs[,"aB"]),stacked.freqs[,"AB"][1]))))
 	centroid<-st_centroid(pol)
	text(centroid[1]/3,centroid[2],col="black","ab",cex=1.4) 
		
	polygon(x=x.polygon,c(stacked.freqs[,"aB"],rev(stacked.freqs[,"ab"])),col="red")
	pol <-st_polygon(list(cbind(c(x.polygon,1) ,c(stacked.freqs[,"aB"],rev(stacked.freqs[,"ab"]),stacked.freqs[,"aB"][1]))))
 	centroid<-st_centroid(pol)
	if(any(stacked.freqs[,"aB"]<0.98)) text(centroid[1],centroid[2],col="black","aB",cex=1.4) 

}
```

##14.3.1 The hitchhiking of neutral alleles, Fig. 14.1

#In this scenario, "A beneficial mutation B arises on the background of a neutral allele whose initial frequency is pA = 10%. The beneficial allele has a strong, additive selection coefficient of hs = 0.05."
```{r}
###Neutral Hitchhiking
n.gens<-500

#B starts at low freq. on the A background
p<-  c(0.001,0.099,0,0.9); names(p)<-c("AB","Ab","aB","ab")

#w.add sets up the relative fitnesses of the four haplotypes. The haplotypes with our beneficial allele "B" have a slight increase in relative fitness (hs = 0.05)
w.add<-c(1,0.95,1,0.95);names(w.add)<-names(p)
#need to set up relative fitnesses of haplotypes differently. How many copies of B allele? add up number and their relative fitnesses.
w.mat<-outer(w.add,w.add,FUN="+")
layout(t(1:3))
par(mar=c(4,4,3,1))

#run function for three different recombination rates, 0.0005, 0.005, 0.05
p.out<-two.loc.sims(p,w.mat,r=0.0005); stack.freqs.plot(p.out,my.title="r=0.0005")
p.out<-two.loc.sims(p,w.mat,r=0.005); stack.freqs.plot(p.out,my.title="r=0.005")
p.out<-two.loc.sims(p,w.mat,r=0.05);
stack.freqs.plot(p.out,my.title="r=0.05")
```
##Based on the above plots, at what level of recombination is the A allele hitchhiking with selection on the B allele? At what level of recombination does the neutral allele no longer hitchhike with selection on the B allele?


##14.3.2 The hitchhiking of deleterious alleles, Fig. 14.2
In this case, our B allele is still beneficial, but now our A allele is no longer neutral, but instead deleterious.
```{r}
par(mar=c(4,4.1,3,1))
n.gens<-1000

#same initial haplotype frequencies as above.
p<-  c(0.001,0.099,0,0.9); names(p)<-c("AB","Ab","aB","ab")

#we have changed our relative fitness such that the ab haplotype has the highest relative fitness. Our B beneficial allele shows up in a haplotype with the deleterious A allele.
w.add<-c(0.97,0.93,1,0.95);names(w.add)<-names(p)
w.mat<-outer(w.add,w.add,FUN="+")

layout(t(1:3))
 p.out<-two.loc.sims(p,w.mat,r=0.00);stack.freqs.plot(p.out,my.title="r=0.0");
 p.out<-two.loc.sims(p,w.mat,r=0.000005);stack.freqs.plot(p.out,my.title="r=0.000005");
 p.out<-two.loc.sims(p,w.mat,r=0.0002);stack.freqs.plot(p.out,my.title="r=0.0002")
```
#What is happening to the Ab haplotype? Why? What is the difference between the three plots in patterns of AB frequency?

#14.3.3: Clonal interference between favourable alleles Fig 14.3. Beneficial alleles A and B arise on different haplotypes.
```{r}
layout(t(1:2))
n.gens<-1000
#There are no AB individuals to begin
p<-  c(0.0,0.01,0.01,0.98); names(p)<-c("AB","Ab","aB","ab")
w.add<-c(1.14,1.08,1.06,1);names(w.add)<-names(p)
w.mat<-outer(w.add,w.add,FUN="+")
p.out<-two.loc.sims(p,w.mat,r=0.00)
stack.freqs.plot(p.out, my.title="r=0.00")
#ignore the AB label on this plot

p<-  c(0.0,0.01,0.01,0.98); names(p)<-c("AB","Ab","aB","ab")
w.add<-c(1.1,1.06,1.055,1);names(w.add)<-names(p)
w.mat<-outer(w.add,w.add,FUN="+")
p.out<-two.loc.sims(p,w.mat,r=0.001)
stack.freqs.plot(p.out, my.title="r=0.001")
```
#What is the difference in the two plots above? Which is the more beneficial allele, A or B? How does that relate to what haplotype will fix with no recombination?


#14.3.4: Epistatic combinations of alleles and the cost of recombination.
"Imagine a pair of alleles A and B at two loci that work very well together, and offer a fitness advantage over the ancestral combination of allele a and b. You could for example imagine that A and B are changes in a protein and its receptor, and that they offer a much more efficient signaling response. However, imagine that A doesn’t work with b, nor does the allele a work well with B."
```{r}
n.gens<-200
layout(t(1:3))
p<-  c(0.3,0.0,0.0,0.7); names(p)<-c("AB","Ab","aB","ab")
w.add<-c(1.05,0.1,0.1,1.0);names(w.add)<-names(p);w.mat<-outer(w.add,w.add,FUN="+")
p.out<-two.loc.sims(p,w.mat,r=0.00)
stack.freqs.plot(p.out, my.title="r=0.00")

p<-  c(0.3,0.0,0.0,0.7); names(p)<-c("AB","Ab","aB","ab")
w.add<-c(1.05,0.1,0.1,1.0);names(w.add)<-names(p);w.mat<-outer(w.add,w.add,FUN="+")
p.out<-two.loc.sims(p,w.mat,r=0.05)
stack.freqs.plot(p.out, my.title="r=0.05")


p<-  c(0.3,0.0,0.0,0.7); names(p)<-c("AB","Ab","aB","ab")
w.add<-c(1.05,0.1,0.1,1.0);names(w.add)<-names(p);w.mat<-outer(w.add,w.add,FUN="+")
p.out<-two.loc.sims(p,w.mat,r=0.1)
stack.freqs.plot(p.out, my.title="r=0.1")
```
